---
output: 
  md_document:
    variant: markdown_github
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# ezsummary 
[![Build Status](https://travis-ci.org/haozhu233/ezsummary.svg?branch=master)](https://travis-ci.org/haozhu233/ezsummary)  
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/ezsummary)](http://cran.r-project.org/package=ezsummary)  
[![Coverage Status](https://coveralls.io/repos/github/haozhu233/ezsummary/badge.svg?branch=master)](https://coveralls.io/github/haozhu233/ezsummary?branch=master)

## Introduction
Hadley's [`dplyr`](https://github.com/hadley/dplyr) provides a grammar to talk about data manipulation and another his package, [`tidyr`](https://github.com/hadley/tidyr) provides a mindset to think about data. These two tools really makes it a lot easier to perform data manipulation today. This package `ezsummary` packed up some commonly used `dplyr` and `tidyr` steps to generate data summarization to help you save some typing time. It also comes with some table decoration tolls that basically allows you to pipe the results directly into a table generating function like `knitr::kable()` to render out.

For example, if you only use `dplyr` and `tidyr` to generate a statistical summary table by group. You need to go through the following steps.
```{r dplyr_tidyr}
library(dplyr)
library(tidyr)

mtcars %>%
  select(cyl, mpg, wt, hp) %>%
  group_by(cyl) %>%
  summarize_each(funs(mean, sd)) %>%
  gather(variable, value, -cyl) %>%
  mutate(value = round(value, 3)) %>%
  separate(variable, into = c("variable", "analysis")) %>%
  spread(analysis, value) %>%
  mutate(variable = factor(variable, levels = c("mpg", "wt", "hp"))) %>%
  arrange(variable, cyl) %>%
  kable()
```

For people who are familar with "tidyverse", I'm sure the above codes are very straightforward. However, it's a bit annoying to type it again and again. With `ezsummary`, you don't need to think too much about it. You can just type:

```{r into_ezsummary}
library(ezsummary)

mtcars %>%
  select(cyl, mpg, wt, hp) %>%
  group_by(cyl) %>%
  ezsummary() %>%
  kable()
```

## To install
``` r
    install.packages("ezsummary")
```

Or  

``` r
    install.packages("devtools")
    devtools::install_github("haozhu233/ezsummary")
```
  
## To use
Here, I will show another quick demo of how to use this package here. For detailed package documentation, please check the [package vignette](http://rpubs.com/haozhu233/ezsummary0_2_0). 

```{r example}
library(dplyr)
library(ezsummary)

mtcars %>%
  # q: quantitative/continuous variables; c: categorical variables
  var_types("qcqqqqqcccc") %>%
  group_by(am) %>%
  ezsummary(flavor = "wide", unit_markup = "[. (.)]",
            digits = 1, p_type = "percent") %>%
  kable(col.names = c("variable", "Manual", "Automatic"))
```

## Issues
If you ever find any issues, please feel free to report it in the issues tracking part on github. [https://github.com/haozhu233/simple.summary/issues](https://github.com/haozhu233/simple.summary/issues). 

Thanks for using this package!
